Algebraic Groups\vspace{-1ex}

### 2020-02-12 Algebraic Geometry/Algebraic-Geometry.tex
Update and fix some old stuff

>>>>

\section{February 5th, 2020}
Today Adam is talking more about limits!

Recall that Yoneda says that the natural transformations 
\[\Hom(\Hom_\calC(-,\lim k),\operatorname{cone}(-,K))\simeq\operatorname{cone}(\lim K,K)\]

\subsection{Equilizers}
Some clarity here: an equilizer in $\Ab$ can be a couple things:
\begin{itemize}
	\item The limit of a pair of parallel functors.
	\item It is the same as saying that the following is exact:
	\[0\to A\to B\xrightarrow{f-g} C\]
\end{itemize}

\subsection{Base change}
This is one of the most important (co)limits in this subject (and in number theory and representation theory).
If $A$ is a $k$-algebra and there is an inclusion $k\hookrightarrow L$. How can we extend $A$ to an $L$-algebra?

Say that $B$ is an extension of $A$ that is also an $L$-algebra. Then if the universe loves us, we should get 
\begin{figure}[h]
	\centering
	\begin{tikzcd}
		k\ar[r]\ar[d] & A\ar[d]\\
		L\ar[r] & B
	\end{tikzcd}
\end{figure}
so if we want the ``best'' one, we need to find the colimit of $L\leftarrow k\to A$. To define this thing we just 
trace through and show that the object with the usual axioms of the tensor product falls out. :)

\begin{prop}
	The following is a pullback (in schemes, locally ringed spaces, etc)
	\begin{figure}[h]
		\centering
		\begin{tikzcd}
			\Spec A\otimes_k L\ar[r]\ar[d] & \Spec L\ar[d]\\
			\Spec A\ar[r] & \Spec k
		\end{tikzcd}
	\end{figure}
\end{prop}
\begin{rmk}
	Why is it called a fiber product? Look at what happens in $\Set$. Look at the limit of the diagram 
	$\ast\stackrel{y}{\hookrightarrow} Y \xleftarrow{f}X$. The pullback is literally the fiber over $y\in Y$.
\end{rmk}
Proving the following is a good excercise. Go try it! It's a fun categorical proof. You can do it by looking at cones. 
\newpage
\begin{lem}
	If $\pi:X\to Y$ is a monomorphism, then in the pullback square 
	\begin{figure}[h]
		\centering
		\begin{tikzcd}
			X'\ar[r]\ar[d,"\pi'"] & X\ar[d,"\pi"]\\
			Y'\ar[r] & Y
		\end{tikzcd}
	\end{figure}

	\noindent $\pi'$ is also mono.
\end{lem}
As a corollary of this, if we have an open $U\subseteq X$ in a scheme and $\pi:Y\to X$ a morphism, then 
$\pi^{-1}(U)$ is open in $Y$.

\subsection{Gluing, baby}
For the setup, remember that $X$ is a topological space and $X=\cup_i U_i$ open cover. We are given sheaves 
$\calF_i$ on $U_i$ together with isomorphisms 
\[b_{ij}:\calF_i|_{U_i\cap U_j}\to\calF_j|_{U_i\cap U_j}.\]

We saw that the right way to glue to an $\calF$ on $X$ is to take 
$\calF(V)$ to be a limit of the diagram we drew yesterday (just the bottom two rows).
Then there are two questions we ask:
\begin{itemize}
	\item Is $\calF|_{U_i}=\calF_i$? This is answered by the \textit{cocycle condition}.
	\item Is $\calF$ a sheaf? This is due to the fact that limits commute with limits.
\end{itemize}

\section{February 7th, 2020}
Max's back. Time to talk about 
\subsection{Separatedness and Properness}
These are replacements for the Hausdorff and Compact conditions, neither of which we get in algebraic geometry. Let's get some definitions 
\begin{defn}
	A morphism $f:X\to Y$ is \textbf{separated} if the diagonal $\Delta:X\to X\times_Y X$ is a closed immersion. This 
	replaces Hausdorff.
\end{defn}
Recall that a closed immersion is a homeomorphism onto a closed subspace such that the map of rings is surjective. 
The prototype you should be thinking of is $\Spec k[x_1,\dots,x_m]/(f_1,\dots,f_n)$ as a closed subspace of $\A m$.
\begin{defn}
	A morphism $f:X\to Y$ is \textbf{closed} if the closed subsets of $X$ are sent to closed subsets of $Y$ (as topological spaces).

	It is \textbf{universally closed} if for all $T\to Y$, the base change map $f_T:X\times_Y T\to T$ is closed.
\end{defn}
\begin{ex}
	Consider $\A1_k\to\Spec k$. This is closed, but not universally closed! Consider the base change via $\A1\to\Spec k$:
	$\A1\times\A1=\A2\to \A1$. In coordinates, we get the map of algebras $k[s]\hookrightarrow k[t,s]$. But the hyperbola 
	cut out by $ts-1$ in $\A2$ projects onto $\A1\setminus 0$ under this map, which is super not closed.

	Is $\A1\to \Spec k$ separated? Hell yes! Look at the ring map $A\times A\to A$ by multiplication. This is surjective.
	And unravelling the definitions gets you the result: every morphism of affines is separated! So we don't really need to check 
	that the inclusion is a topological embedding onto a closed set.
\end{ex}

I (well, Max) claim(s): If $f:X\to Y$ is separated and $\sigma:Y\to X$ is a section, then $\sigma$ is closed.
In fact, check this out: \textbf{any section is a pullback of the diagonal} or even \textbf{``The diagonal is the universal section.''}

\begin{figure}[h]
	\centering
	\begin{tikzcd}
		X\ar[r]\ar[d,bend right,"f",swap]\ar[rr,bend left,"\id"] & X\times_Y \tilde X\ar[r]\ar[d,swap,"f_X"] & \tilde X\ar[d,"f"]\\
		Y\ar[u,"\sigma",swap] \ar[r,"\sigma"] & X\ar[u,bend right,swap,"\Delta"] \ar[r,"f"] & Y
	\end{tikzcd}
\end{figure}
Check it out. Assuming everything is a pullback square (remember the neat two of three thing for pullbacks) we get that $\Delta$ is 
a pullback through itself of the diagonal! This gives us another result: $\Delta$ is closed if and only if sections are universally closed.

So sections are closed, but ot universally closed! For instance the line with two origins is not separated!

\begin{defn}
	A morphism $f:X\to Y$ is \textbf{proper} if it is of finite type, separated, and universally closed.
\end{defn}
Note: $f$ is separated if and only if ``sections of $f$'' are universally closed. An example, $\P n$ is proper over $\Spec\bbZ$.
So of course it is proper over anything.

\section{February 10th, 2020}
Today Taffy is talking about base change!

\subsection{Base change}
Notice this cool thing that we see in topological spaces: let $f:U\hookrightarrow Y$ be an open embedding. Then consider the pullback
\begin{figure}[h]
	\centering
	\begin{tikzcd}
		X\times_Y U\ar[r]\ar[d,"g"] & U\ar[d,"f"]\\
		X\ar[r] & Y
	\end{tikzcd}
\end{figure}
where we see $X\times_Y U\cong f^{-1}(U)$ and $g:f^{-1}(U)\hookrightarrow X$ is the open embedding. Here we just did base change and we saw a property being preserved!

In $\Sch$, we can do the same kind of thing: Let $\calX$ and $\calY$ be schemes over $k$.

I just listened to the rest. Good talk!

\section{February 12th, 2020}
Today we are going to talk more about seperatedness and properness and in particular the so-called 
\subsection{Valuative Criteria}
The idea here is that checking for separatedness or properness can be quite daunting sometimes. We will give a way to do this!
\begin{defn}
	A domain $R$ is a \textbf{valuation ring} if for every $x\in \kappa(R)^\times$, we have $x\in R$ or $x^{-1}\in R$.
\end{defn}
\begin{lem}
	Consider loal rings contained in a field $k$ (with fraction field $k$). Say that $(S,\frakm_S)$ \textit{dominates} $(R,\frakm_R)$ if $R\subset S$ 
	and furthermore $\frakm_S\cap R=\frakm_R$. This gives a partial ordering.

	Then the valuation rings in $k$ are precisely the maximal elements for this ordering.
\end{lem}
\begin{defn}
	A morphism $f:X\to Y$ is \textbf{quasi-separated} if the diagonal $\Delta:X\to X\times_Y X$ is quasicompact.
\end{defn}
THis is always true if $X$ is locally Noetherian. This will be true in almost every case you see in real life, because it 
is really necessary to make the entire theory work.

\begin{defn}
	A \textbf{test diagram} for a morphism $f:X\to Y$ is a diagram as in the figure below	
	where $i:R\hookrightarrow k=\kappa(R)$ is a vaulation ring.
	A \textbf{filling} for a test diagram is an arrow $\gamma:\Spec R\to X$ making this diagram commute.
\end{defn}
\begin{figure}[h]\label{fig:testdiagram}
	\centering
	\begin{tikzcd}
		\Spec k\ar[r]\ar[d,"i^\ast"] & X\ar[d,"f"]\\
		\Spec R\ar[r] & Y
	\end{tikzcd}
	\caption{A test diagram for $f$}
\end{figure}
\begin{thm}
	\begin{enumerate}
		\item A quasiseparated morphism $f:X\to Y$ is separated if and only if every test diagram has at most one filling.
		\item A quasicompact morphism $f:X\to Y$ is universally closed if and only if every test diagram admits at least one filling.
		\item A quasisemparated morphism of finite type $f:X\to Y$ is proper sif and only if every test digram admints precisely one filling.
	\end{enumerate}
\end{thm}
\begin{lem}
	The entire theorem above follows from (b).
\end{lem}
\begin{prf}
	To get (a) from (b), $f$ is quasiseparated iff $\Delta:X\to X\times_YX$ is quasicompact. Note that $f$ is separated iff $\Delta$ is 
	universally closed. Further, note that $\Delta$ is always an immersion. One can show that $\Delta$ is a closed immersion into an open subscheme (look in Hartshorne and use a cover by open affines).

	But then the test diagram for $f$ with two fillings $\alpha,\beta$ gives us a test diagram for $\Delta$ with $(\alpha,\beta)$ along the bottom.
	What does it mean for this new diagram to have a filling? If we ahve $t:\Spec R\to X$, we get that 
	\[(\alpha,\beta)=\Delta\circ t=(t,t)\]
	so we have a filling if and only if $\alpha=\beta$!

	That (b) implies (c) follows from the definitions!
\end{prf}
\begin{defn}
	Given a scheme $X$, a point $s\in X$ is a \textbf{specialization} of a point $t\in X$ (equivalently, $t$ 
	is a generization of $s$) if $s\in\overline{\{t\}}$.
\end{defn}
This is a topological property, but we want to capture it algebraically. We need to pass to limiting algebraic object 
to do it most efficiently. As a mater of notation, we write $t\rightsquigarrow s$.
\begin{lem}
	Suppose $t\rightsquigarrow s$ in $X$ and we are given $K\supset \kappa(t)$ (residue field). Then there exists a vaulation ring $R\subset K=\operatorname{Frac} R$ and 
	morphism $\Spec R\to X$ such that $(0)\mapsto t$ and $\frakm_R\mapsto s$.
\end{lem}
A ``proof'' of this is to look at $\O_{\overline{\{t\}},s}\subset k$ and find $R$ using Zorn's lemma with respect to domination.

### 2020-02-12 Algebraic Groups/Algebraic-Groups.tex
Update and fix some old stuff

>>>>

\section{February 5th, 2020}
Today's plan: we're going to go back and talk a bit more about the Frobenius twist to understand it better.
Then we'll talk about restricted enveloping algebras and then do some synthesis.

\subsection{Frobenius twist}
Let's saw we have a commutative ring with $R$ modules $A$ and $B$. Then we can take the pushout to get $A\otimes_R B$. Instead, let's
take a $k$-algebra and an automorphism $\varphi:k\to k$ we take the pushout 
\begin{figure}[h]
	\centering
	\begin{tikzcd}
		k\ar[r]\ar[d,"\varphi"] & A\ar[d]\\
		k\ar[r] & A\otimes_\varphi k
	\end{tikzcd}
\end{figure}
which is the twist of $A$ by the automorphism $\varphi$. Here the scalars are controlled by $\varphi$ since 
\begin{align*}
	\lambda(a\otimes \mu)&= a\otimes \lambda\mu\\
	&=\varphi^{-1}(\lambda)a\otimes \mu
\end{align*}
whenever this all makes sense.

Now letting $\varphi(x)=x^p$ we get the Frobenius twist $A\otimes_\varphi k\eqdef A^{(1)}$. Whenever $k$ is perfect, this means that if $A$ is an 
abelian group, so is $A^{(1)}$.

We always have the map $a^{(1)}\eqdef a\otimes 1\to a^p$ end extending linearly to the map 
\[a\otimes\lambda=\lambda^{1/p}a\otimes 1\mapsto \lambda a^p.\]
\begin{rmk}
	While this gives us a map $A^{(1)}\to A$, we don't ge a map in the other direction or $M^{(1)}\to M$ for a vector space $M$.
	But we do get a map $M^{(1)}\to S^p(M)$.
\end{rmk}
\begin{rmk}
	Let $X/k$ be a scheme. Then we can define $X^{(1)}/k$. If $X=G$ is a group scheme, then if $A=k[G]$,
	we can define $G^{(1)}$ to be the group scheme such that 
	\[k[G^{(1)}]=k[G]^{(1)}\]
\end{rmk}

You can read more about these in Jantzen, but the map $A^{(1)}\to A$ gives us a map $G\to G^{(1)}$
and the kernel of this map is the Frobenius kernel.

\subsection{Representation theory}
\begin{rmk}
	If $M\in\Rep G$, then this induces a representation of $\Lie G$. Why is this true? Well we can use the functoriality 
	of $\Lie$! In particular, a representation $M$ of $G$ is equipped with a morphism $G\to \GL(M)$, yielding a lie algebra morphism 
	\[\Lie G\to \Lie GL(M)=\gl(M)\]
\end{rmk}

The Frobenius twist also gives us maps on the level of representations. That is, we get a functor 
\[(-)^{(1)}:\Rep G\to \Rep G\]
More clearly, if $M\in\Rep G$ we have this vector space as well as a coaction 
\[k[G]\to M\otimes k[G].\]
This gives us a map 
\[M^{(1)}\to M^{(1)}\otimes k[G]^{(1)}\to M^{(1)}\otimes k[G]\]
so $M\in \operatorname{comod}k[G]\simeq \Rep G$.

If we have two representations $M,N\in\Rep G$, then the map 
\[\Hom(M,N)\to \Hom(M^{(1)},N^{(1)})\]
Considering $\End(M)\to \End(M^{(1)})$, this gives us a map $\GL(M)\to\GL(M^{(1)})$.

\begin{rmk}
	The functor $(-)^{(1)}$ is what is called a ``strict polynomial functor''. :)
\end{rmk}

\begin{ex}
	${\Ga}_{(1)}\simeq \alpha_p$ is the group represented by $k[t]/t^p$ with $t$ primitive. ${\Gm}_{(1)}$
	is the same algebra with $t$ grouplike.
\end{ex}
\begin{ex}
	We have already seen the coordinate algebra of $\GL_n$. What is the Frobenius twist? It is taking a matrix 
	where we take all entries to the $p^{th}$ power (not the whole matrix!). So then the kernel is 
	\[{\GL_n}_{(1)}=\{(a_{ij})|a_{ij}^p=\delta_{ij}\}\]

	If you try to evaluate at a field, you won't get much: only one point. Thus this is a highly-singluar connected, finite 
	group scheme with coordinate algebra 
	\[k[x_{ij}]/(x_{ij}^p-\delta_{ij})\]

	This gives you the ``infinitesimal neighborhood of 1''.
\end{ex}

\subsection{Restricted Lie algebra}
\begin{rmk}
	We never mentioned what happens for $\gl_n$! Here we can honestly multiply matrices, so we can define $A^{[p]}=A^p$. This tells you that for 
	matrix groups we can realize their power operation in this way, but of course this is undesirable as it is dependent on embedding.
\end{rmk}

Recall that the universal enveloping algebra of a Lie algebra $\frakg$ is 
\[\calU(\frakg)=T(\frakg)/\langle x\otimes y-y\otimes x-[x,y]| x,y\in\frakg)\]
\begin{defn}
	If $\frakg$ is a restricted Lie algebra (say $\Lie G$), then the \textbf{restricted Lie algebra} is 
	\[u(\frakg)=\calU(\frakg)/\langle x^p-x^{[p]}|x\in\frakg)\]
\end{defn}
We have PBW bases for both $\calU(\frakg)$ and $u(\frakg)$. The latter is better because if $x_1,\dots,x_n$ are a basis for $\frakg$,
the basis is 
\[\{x_1^{a_1}\cdots x_n^{a_n}|a_i\le p-1\}.\]

One can show that the ideal generated by $x^p-x^{[p]}$ is a Hopf ideal, so one gets that $u(\frakg)$ is again a cocommutative Hopf algebra, 
but this time it is finite dimensional! Then one can consider just the $p$-restricted representations of $\frakg$ and one 
finds that this category is equivalent to $\lmod {u(\frakg)}$.
\begin{ex}
	If $\frakg_m=\Lie\Gm$, we can compute 
	\[u(\frakg_m)=k[t]/(t^p-1)\]
	but now $t$ is primitive! In fact, $u(\frakg_m)=k[\mu_p]^\ast=k[\bbZ/p]$.
\end{ex}
The above example highlights an important result: if $\frakg=\Lie G$,
\[u(\frakg)\xrightarrow{\sim}k[G_{(1)}]^\ast\]
and so we get $\Rep\frakg\simeq \Rep G_{(1)}$. Recall in manifolds that the theories are almost of Lie groups and algebras 
are very tightly connected (for connected compact manifolds). Here we don't get that! However we get a shadow of it:

Note that we actually get a nested sequence of Frobenius kernels:
\[G\supset G_{(n)}\supset\cdots\supset G_{(1)}\]
meaning that in this way we get better and better approximations of $G$.

\section{February 7th, 2020}
Today we are going to talk about induction and restriction.
\subsection{The fixed points functor}
Recall the \textit{fixed point functor}: let $G$ be an algebraic group scheme and $M\in\Rep G$ (which we identify with $\lmod G$).
Given an action of $G$ on $M$, we get 
\[M^G=\{m\in M|\forall R, g\in G(R), g\cdot(m\otimes 1_R)=m\otimes 1_R\}\]
This definition becomes difficult because it requires an extension of scalars and when you have a lot of nilpotents it messes with things.

An alternative method: define 
\[M^G=\{m\in M|\Delta_M(m)=m\otimes 1\}\]
where $\Delta_M:M\to M\otimes k[G]$ is the coaction map. We can also write 
\[M^G=\ker(\Delta_M-\id_M\otimes 1).\]

This does indeed create a functor $(-)^G:\Rep G\to \Vectk$.
\begin{prop}
	$(-)^G$ is left exact.
\end{prop}
\begin{rmk}
	$\Rep G$ has an internal Hom. That is, given $M,M'\in\Rep G$, we have an action of $G$ on $\Hom_k(M,M')$ (note: note $G$-equivariant maps!) via the diagonal (in the usual sense of 
	Hopf algebras). All of this, together with tensor-(internal) hom adjunction
	\[\Hom_A(M\otimes_k N, M')\cong\Hom_G(M,,\Hom_k(N,M')),\]
	gives the category of Hopf modules the structure of a rigid monoidal category.

	But now (as vector spaces), we have the isomorphism 
	\[\Hom_k(M,M')^A\cong \Hom_A(M,M').\]
\end{rmk}

Let's take a moment to consider $k[G]$ as a $G$-module. The \textbf{right regular representation} on $k[G]$
is as follows: given $f\in k[G]$, we have 
\[\rho_r(g)(f)(x)=(g\cdot f)(x)=f(x\cdot g)\]
and 
\[\rho_l(g)(f)(x)=(g\ast f)(x)=f(g^{-1}\cdot x)\]
and by composing the two, we get $\rho_r\circ\rho_l=\rho_l\circ\rho_g$ to get the adjoint representation on $k[G]$.

Now as comodules, we have the action 
\[\Delta:k[G]\to k[G]\otimes k[G]\]
which is right regular representation. \textbf{Why?} We can define the left regular representation by 
\[\tau\circ(S\otimes \id)\circ\Delta.\]
As an exercise, determine whether this is the same element as $(\id\otimes S)\circ \Delta$.

\subsection{Induction and restriction}
Given $H\hookrightarrow G$, this gives us an obvious functor 
\[\Res_H^G:\Rep G\to \Rep H\]
Then we claim (but don't prove here) that $\Res$ has a right adjoint. In some cases, there is a left adjoint as well although 
we don't have that in general.

In this case, with groups, we can define 
\[\Ind_H^G M\eqdef (M\otimes_kk[G])^H\]
where each are $H$-modules ($k[g]$ with the right regular action) and the action on the tensor is via the left regular action.

Another point of view: remember we can define the affine group scheme $M_a$ where $M_a(X)=M\otimes X$. Then 
\[M\otimes k[G]=M_a(k[G])\]
and then we can look at the morphisms $G\to M_a$ and consider 
$(M\otimes k[G])^H$ to be the \textit{$H$-invariant functions $f:G\to M$} where $H$ actions on $\Mor(G,M_a)$ via 
\[(h\cdot f)(g)=hf(gh)\]
and so 
\[\Mor(G,M_a)^H=\{f:G\to M|f(gh)=h^{-1}f(g),\forall R, g\in G(R)\}.\]

\begin{rmk}
	Notice that 
	\[\Ind_H^G=(-)^H\circ(-\otimes k[G])\]
	and so (since tensor is exact and restriction is left exact), $\Ind_H^G$ is left exact.
\end{rmk}

\subsection{Frobenius reciprocity}
Define the evaluation map $\varepsilon_M:\Ind_H^G M\to M$ via 
\[m\otimes f\mapsto \varepsilon(f)m\]
Using the other form, if we have $f\in\Mor(G,M)^H$, we get $f\mapsto f(1)\in M$.

\begin{prop}\label{prop:frobenius}
	\begin{enumerate}
		\item $\varepsilon_M:\Ind_H^G M\to M$ is an $H$-module map.
		\item (\textit{Frobenius reciprocity}) $\varepsilon_M$ induces an isomorphism 
		\[\Hom_G(N,\Ind_H^G M)\xrightarrow{\sim}\Hom_H(\Res_H^G N, M).\]
	\end{enumerate}
\end{prop}
\begin{cor}
	\begin{enumerate}
		\item $\Ind$ is transitive: $\Ind_H^G=\Ind_{H'}^G\circ\Ind_H^{H'}$.
		\item $\Ind$ commutes with field extensions.
		\item $\Ind$ preserves injectives.
	\end{enumerate}
\end{cor}
\begin{rmk}
	Recall the proof that $\lmod R$ has enough projectives: the idea was to construct projective in $\Ab$ and then 
	use a kind of induction to bring them up to the module category. In general, property (c) above empowers us to show 
	that categories have enough injectives!
\end{rmk}
We have a week off! So here's some homework:
\begin{prob}
	Prove proposition \ref{prop:frobenius}. Notice if we have 
	\[N\xrightarrow{f}\Ind_H^G\xrightarrow{\varepsilon_M} M\]
	and define $\tilde\Psi^H:N\to \Ind_H^G M$ via the $H$-map $\psi:N\to M$ where 
	\[\tilde\Psi:N\to \Mor(G,M)\]
	via $n\mapsto \tilde\Psi_n:G\to M$ where $\tilde\Psi_n(g)=\Psi(g^{-1}n)$. Then show $\tilde\Psi_n$ is $H$-invariant, a $G$-map, and prove there are mutual inverses.
\end{prob}
